Notes on Validated Model Counting \\ Version of \today

### 2022-04-11 doc/mc-notes.tex
Ready for review

{\bf Example 1}: Let $\ring$ be the set of rational numbers and
$\cost(x_i) = 1/2$ for all variables $x_i$.  The cost of every
assignment will be $1/2^{n}$, and the cost of a function will be its
{\em density}, denoted $\density(f)$.  That is $\density(f) \in [0,
  1]$ will be the fraction of assignments for which the function
evaluates to 1.  Given an ability to compute the density of function
$f$, we can scale this by $2^n$ to compute the total number of models
$|\interpset{f}|$.  This is the core task of model counting.  Using
density as our metric has the advantage that it does not vary when the
function is embedded in a larger domain $X' \supseteq X$.  As a
variant, we can support some forms of weighted model counting by
assigning weights of the form $\cost(x_i) = w_i$, where $0 \leq w_i
\leq 1$ for every variable $x_i$.

@@

We introduce the CRAT clausal proof framework for generating a
checkable proof of the valuation of a Boolean formula, given in
conjunctive normal form, according to a specified cost function.  A
CRAT proof provides two capabilities: 1) a {\em schema} for computing a cost
function using the ring operations, and 2) a {\em proof} that this schema
accurately characterizes the cost function for the input formula.
A schema can be either {\em function independent}, meaning that it holds
for any cost function, or {\em function dependent}, meaning that
it holds for only for a specific cost function.

The CRAT format draws its inspiration from the QRAT format for
quantified Boolean formulas (QBF).  The following are its key properties:
\begin{itemize}
\item Clause addition is allowed, as long as either 1) the new clause is
  blocked, or 2) it satisfies the RUP/RAT property with respect to a sequence
  of earlier antecedent clauses.
\item  Extension variables can be introduced only according to the operations $\pand$, $\por$, and {\em abstraction}
\begin{itemize}
\item The checker tracks the dependency set for every input and
  extension variable.  When an extension variable is introduced based
  on the $\pand$ operation, the dependency sets of its arguments must
  be disjoint.
\item When an extension variable is introduced based on the $\por$
  operation, the step must cite earlier steps providing a RUP/RAT proof
  that the two arguments are mutually exclusive.
\item The {\em abstraction} operation allows an extension
  variable to serve as an alias for one of more other 
  variables.  Use of this rule renders the schema function dependent.
\item Boolean complement is provided implicitly by allowing the
  arguments of the extension operations to be literals and not just
  variables.
\end{itemize}
\item Unlike DRAT, the deletion of a clause requires showing that this clause is implied by one or more other clauses.
\item Unlike QRAT, it need not support universal quantification.
\end{itemize}

@@

A CRAT proof follows the same general form as a QRAT dual proof---one
that ensures that each clause addition and each clause deletion preserves
equivalence.  Starting with the set of input clauses, it produces a
sequence of steps that both add and delete clauses.  Each addition
must be truth preserving and each deletion must be falsehood
prserving.  At the end, all input clauses must have been deleted, and
among the remaining clauses there must be only a single unit clause consisting of some variable or
its complement.  Except for trivial cases, the final variable will be
an extension variable.  The sequence of extension operations define
the schema for computing the cost function.  That is, a value for the
cost function is computed for every literal according to the rules of
Lemmas \ref{lemma:complementation}--\ref{lemma:disjunction}.  For
every abstraction step, all literals having the indicated extension
variable as their alias must evaluate to the same value.  The computed
cost is then the value computed for the literal in the final unit
clause.

@@

\begin{figure}
  \centering{
    \begin{minipage}{0.4\textwidth}
    A) Logical Schema\\[2ex]
    \centering{\input{dd/c3}}
    \end{minipage}
    \begin{minipage}{0.4\textwidth}
    B) Evaluation Schema\\[2ex]
    \centering{\input{dd/c3-eval}}
    \end{minipage}
  }
\caption{Schema for Formula $x_1 \lor x_2 \lor x_3$}
\label{fig:c3:schema}
\end{figure}

@@

As an illustration, consider the Boolean formula $x_1 \lor x_2 \lor
x_3$, represented by a single clause.  We cannot directly use the
$\por$ operation to form this disjunction, since the sets of assignments
satisfying the individual literals are not disjoint.  Instead, we must
decompose this formula into a sequence of operations, as shown by the
logical schema of Figure~\ref{fig:c3:schema}A\@.  Circles in this schema
correspond to $\pand$ operations and squares to $\por$ operations.
The subscripts of the variables and the operator labels correspond to
the numbers of the input and extension variables in the CRAT proof.

@@

Figure~\ref{fig:c3:crat} shows an annotated version of the CRAT proof
for this example.  Clause \#1 corresponds to the input formula, and
clauses \#2--\#13 are the defining clauses for the four operations.
Clause \#13 adds the unit clause corresponding to sum $s_7$,
indicating that the extension variable variable will evaluate to
$\tautology$.  for any assigment that satisfies the input clause.  We
can write this as $C \turnstile  s_7$ for input clause $C$.  The
deletion step at the end turns this around, showing that
$s_7 \turnstile C_I$, and therefore the input clause can be deleted,
This completes a proof that extension variable $s_7$ is logically
equivalent to the input formula.

\section{What's known, and what's not}

Given an arbitrary CNF formula, we can use BDD operations to generate
a schematic representation.  The proof generation can follow the
methods we have used for generating unsatisfiability proofs of Boolean
formulas~\cite{bryant:tacas:2021} and dual proofs of quantified
Boolean formulas~\cite{bryant:cade:2021}.  The key idea of that work
was to use extended resolution to encode the semantics of the BDD
nodes as part of the proof.  Here we can further decompose each $\ite$
(short for ``if-then-else'') operation representing a BDD node into
two $\pand$ and one $\por$ operation.  That is, we can encode
$\ite(x,A,B$ as $(x \pand A) \por (\obar{x} \pand B$, where $x$ is an
input variable and $A$ and $B$ are subformulas.  In doing so, we must
make sure that $x$ is not in the dependency set of either $A$ or $B$.
For ordered BDDs,~\cite{Bryant:1986} this property holds, since the
variables in $A$ and $B$ must be greater in the variable ordering than
$x$.

The BDD representations of many of the formulas occuring in
model-counting problems are far to large for this approach to be
practical.  One refinement of the approach, implemented by the ADDMC
model counter, is to abstract subformulas, keeping track only of the
number of models they can have, representing a number of subformulas
with a single ADD leaf node.  Our abstraction operation is intended to
support this capability.  Under controlled conditions, we can
introduce a new variable the represents the cost function for some
subformula.  If subsequent formulas are guaranteed to yield the same
cost, then these can alias to the earlier variable.  However, it is
not clear how to prove that this aliasing preserves equivalence.

@@

>>>>

{\bf Example 1}: Let $\ring$ be the set of rational numbers and
$\cost(x_i) = 1/2$ for all variables $x_i$.  The cost of every
assignment is then $1/2^{n}$, and the cost of a function is its
{\em density}, denoted $\density(f)$.  That is, the density of $f$ satisfies
$0 \leq \density(f) \leq 1$.  It
is the fraction of assignments for which the $f$
evaluates to 1, with $\density(\zero) = 0$ and $\density(\one) = 1$.  The density of a function
$f$ can be scaled by $2^n$ to compute the total number of models
$|\interpset{f}|$.  This is the core task of model counting.  Using
density as the metric, rather than the number of models, has the advantage that it does not vary when the
function is embedded in a larger domain $X' \supseteq X$.  As a
variant, some other forms of weighted model counting can be supported by
assigning weights of the form $\cost(x_i) = w_i$, where $0 \leq w_i
\leq 1$ for every variable $x_i$.

@@

We introduce the CRAT clausal proof framework for generating a
checkable proof of the valuation of a Boolean formula, given in
conjunctive normal form, according to a specified cost function.  A
CRAT proof provides two forms of information: 1) a {\em schema} for computing a cost
function using the ring operations, and 2) a {\em proof} that this schema
accurately characterizes the cost function for the input formula.
A schema can be either {\em function independent}, meaning that it holds
for any cost function, or {\em function dependent}, meaning that
it holds for only for a specific cost function.

The CRAT format draws its inspiration from the LRAT format for Boolean formulas and the
QRAT format for
quantified Boolean formulas (QBF).  The following are its key properties:
\begin{itemize}
\item As with LRAT, a clause can be added as long as either 1) it is
  blocked, or 2) it satisfies the RUP/RAT property with respect to a supplied sequence
  of earlier antecedent clauses.
\item  Extension variables can be introduced only according to the operations $\pand$, $\por$, and to an abstraction rule.
\begin{itemize}
\item The checker tracks the dependency set for every input and
  extension variable.  When an extension variable is introduced based
  on the $\pand$ operation, the dependency sets of its arguments must
  be disjoint.
\item When an extension variable is introduced based on the $\por$
  operation, the step must cite earlier steps providing a RUP/RAT proof
  that the two arguments are mutually exclusive.
\item The {\em abstraction} operation allows an extension
  variable to serve as an alias for one of more other 
  variables.  Use of this rule renders the schema function dependent.
\item Boolean complement is provided implicitly by allowing the
  arguments of the extension operations to be literals and not just
  variables.
\end{itemize}
\item Unlike LRAT, the deletion of a clause requires showing that this clause is implied by one or more other clauses.
\item Unlike QRAT, it need not support universal quantification.
\end{itemize}

@@

A CRAT proof follows the same general form as a QRAT dual proof---one
that ensures that each clause addition and each clause deletion preserves
equivalence.  Starting with the set of input clauses, it produces a
sequence of steps that both add and delete clauses.  Each addition
must be truth preserving and each deletion must be falsehood
preserving.  At the end, all input clauses must have been deleted, and
among the remaining clauses there must be only a single unit clause consisting of some variable or
its complement.  Except for trivial cases, the final variable will be
an extension variable.  The sequence of extension operations define
the schema for computing the cost function.  That is, a value for the
cost function is computed for every literal according to the rules of
Lemmas \ref{lemma:complementation}--\ref{lemma:disjunction}.  For
every abstraction step, all literals having the indicated extension
variable as their alias must evaluate to the same value.  The computed
cost is then the value computed for the literal in the final unit
clause.

@@

\begin{figure}
  \centering{
    \begin{minipage}{0.4\textwidth}
      A) Logical Form   \\[2ex]
    \centering{\input{dd/c3}}
    \end{minipage}
    \begin{minipage}{0.4\textwidth}
    B) Evaluation Form\\[2ex]
    \centering{\input{dd/c3-eval}}
    \end{minipage}
  }
\caption{Schema for Formula $x_1 \lor x_2 \lor x_3$}
\label{fig:c3:schema}
\end{figure}

@@

As an illustration, consider the Boolean formula $x_1 \lor x_2 \lor
x_3$, represented by a single clause.  We cannot directly use the
$\por$ operation to form these disjunctions, since the sets of assignments
satisfying the individual literals are not disjoint.  Instead, we must
decompose this formula into a sequence of operations, as shown by the
logical form of Figure~\ref{fig:c3:schema}A\@.  Circles in this schema
correspond to $\pand$ operations and squares to $\por$ operations.
The subscripts of the variables and the operator labels correspond to
the numbers of the input and extension variables in the CRAT proof.

@@

Figure~\ref{fig:c3:crat} shows an annotated version of the CRAT proof
for this example.  Clause \#1 corresponds to the input formula, and
clauses \#2--\#13 are the defining clauses for the four operations.
Each of the two sum operations lists one of the earlier defining
clauses as a proof that its arguments are mutually exclusive.
Clause \#14 adds the unit clause corresponding to sum $s_7$,
indicating that the extension variable variable will evaluate to
$\tautology$  for any assigment that satisfies the input clause.  We
can write this as $C \turnstile s_7$ for input clause $C$.  The
deletion step at the end turns this around, showing that $s_7
\turnstile C_I$, and therefore the input clause can be deleted, This
completes a proof that extension variable $s_7$ is logically
equivalent to the input formula.

\section{Looking Ahead}

\subsection{Implementing Certified Counters}

Given an arbitrary CNF formula, we can use BDD operations to generate
a schematic representation.  The proof generation can follow the
methods we have used for generating unsatisfiability proofs of Boolean
formulas~\cite{bryant:tacas:2021} and dual proofs of quantified
Boolean formulas~\cite{bryant:cade:2021}.  The key idea 
is to use extended resolution to encode the semantics of the BDD
nodes as part of the proof.  Here we can further decompose each $\ite$
(short for ``if-then-else'') operation representing a BDD node into
two $\pand$ and one $\por$ operation.  That is, we can encode
$\ite(x,A,B)$ as $(x \pand A) \por (\obar{x} \pand B)$, where $x$ is an
input variable and $A$ and $B$ are subformulas.  The sum operation trivially satisfies the disjoint assignment requirement, since $x$ has positive polarity in the first product and negative polarity in the second.
We must also
make sure that $x$ is not in the dependency set of either $A$ or $B$.
For ordered BDDs,~\cite{Bryant:1986} this property holds, because the
variables in $A$ and $B$ will be greater in the variable ordering than
$x$.

The BDD representations of many of the formulas occuring in
model-counting problems are far to large for this approach to be
practical.  One refinement of the approach, implemented by the ADDMC
model counter, is to abstract subformulas, keeping track only of the
number of models they can have and representing a number of subformulas
with a single ADD leaf node.  Our abstraction operation is intended to
support this capability.  We can
introduce a new variable to represent the cost function for some
subformula.  If subsequent formulas are guaranteed to yield the same
cost, then these can alias to the earlier variable.  However, it is
not clear how to prove that this aliasing preserves equivalence.

@@

\subsection{TO-DO List}

\begin{itemize}
\item Proof Framework
  \begin{itemize}
  \item Generalities and details of the format
  \item What should be the final state?
  \item When can blocked clauses be deleted?
  \item How can abstraction be incorporated?
  \end{itemize}

\item Checker
  \begin{itemize}
  \item Working prototype
  \item C/C++ (or Rust?)
  \item Formally verified
  \end{itemize}

\item Counters
\begin{itemize}
\item BDD-based
  \begin{itemize}
  \item Prototype
  \item C/C++
  \end{itemize}

\item CDCL-based
  \begin{itemize}
  \item Prototype
  \item C/C++
  \end{itemize}
\end{itemize}

\end{itemize}
